---
title: "EDA_Main_Project"
author: "Siddartha Rao, Shreya Paul and Vishal Singh"
date: "December 14, 2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
library(tidyverse)
library(dplyr)
library(gapminder)
library(tidyverse)
library(lme4)
library(knitr)
library(kableExtra)
library(arm)
#library(ggpubr)
library(viridis)
library(mgcv)
library(broom)
library(RColorBrewer)
cb_palette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
palette(brewer.pal(n = 8, name = "Spectral"))
```

```{r include=FALSE, echo=FALSE}
# Loading Data
spotify.df = read.csv("spotify.dataset.final.csv")
print(paste("Unique Genres:", unique(spotify.df$genre)))
print(paste("Column Names (Features):", colnames(spotify.df)))
```

### Analysis

We began by looking at a few of the basic graphs for our data to get a better sense of our dependent variable 'Popularity' since its is a very subjective term. We look at the most popular and least popular songs from out dataset.

#### Most Popular Songs

```{r}
spotify.df.filtered.popularity = spotify.df %>% filter(popularity <= 100) %>% filter(track_name != 'In My Feelings')
top_n(spotify.df.filtered.popularity, n = 10, spotify.df.filtered.popularity$popularity) %>% ggplot(., aes(x = track_name, y = popularity)) + geom_bar(stat = "identity", fill = "blue") + coord_flip() + labs(x = "Track Name", y = "Popularity", title = "Most popular songs")
```

#### Least Popular Songs

```{r}
top_n(spotify.df.filtered.popularity, n = -10, spotify.df.filtered.popularity$popularity) %>% ggplot(., aes(x = track_name, y = popularity)) + geom_bar(stat = "identity", fill = "red") + coord_flip()  + labs(x = "Track Name", y = "Popularity", title = "Least popular songs")
```

#### Probability Distribution

```{r}
library(plyr)
mu <- ddply(spotify.df, "genre", summarise, grp.mean=mean(popularity))

a = ggplot(spotify.df, aes(x = popularity, color = genre))

a + geom_density(size = 1) + labs(x = "Popularity", y = "Density", title="Density of Popularity for different Genres") + scale_color_brewer(palette = "Spectral") + geom_vline(data=mu, aes(xintercept=grp.mean, color=genre), linetype="dashed", size = 1)
```

#### Mean Popularity by Genre

```{r}
a = spotify.df %>% group_by(genre) %>% dplyr::summarise(cnt = n()) %>% 
  ggplot(aes(x = genre, y = cnt, color = genre)) 
a + geom_bar(stat = "identity", fill = "white", size = 1) + scale_color_brewer(palette = "Spectral") + labs(x = "Genres", y = "Song Count", title="Number of songs for different Genre")#+
```

#### Boxplot of popularity by Genre

```{r}
ggplot(data = spotify.df, aes(x = genre, y = popularity, color = genre)) + geom_boxplot(size=1) + scale_color_brewer(palette = "Spectral") + labs(x = "Genres", y = "Popularity", title="Distribution of Popularity for different Genre")
```



```{r}
#Removing variables track name and ones which have been convertd to categorical
spotify.model.df = spotify.df %>% dplyr::select(-duration_ms, -loudness, -speechiness, -liveness, -tempo, -track_name)
```

#### Popularity by Valence for different Genres

```{r}
ggplot(aes(x = valence, y = popularity, color = genre), data = spotify.model.df) + geom_smooth(method = "lm", se = FALSE) + labs(x = "Valence", y = "Popularity", title="Popularity by Valence for different Genres")
```

#### Popularity by Acousticness for different Genres

```{r}
ggplot(aes(x = acousticness, y = popularity, color = genre), data = spotify.model.df) + geom_smooth(method = "lm", se = FALSE) + labs(x = "Acousticness", y = "Popularity", title="Popularity by Acousticness for different Genres")
```
#### Popularity by Speechiness for different Genres

```{r}
ggplot(aes(x = speechiness, y = popularity, color = genre), data = spotify.df) + geom_smooth(method = "lm", se = FALSE) + labs(x = "Speechiness", y = "Popularity", title="Popularity by Speechiness for different Genres")
```

### Linear model

$Prediction = lm(duration*(loudness + liveness + Tempo + Dancibility) + Valence + Genre + Acousticness + Dancibilty + mode + speechiness)$

```{r}
spotify.lm = lm(popularity~duration_s*(loudness.level+liveness.level+tempo.level+danceability)+valence+genre+acousticness+danceability+mode+speechiness.level,data =spotify.model.df )
summary(spotify.lm)
```

```{r}
### Predicted values
spotify.model.df$predictions = predict(spotify.lm, spotify.model.df[,-spotify.model.df$popularity])

spotify.lm.df = augment(spotify.lm)

ggplot(spotify.lm.df, aes(x = duration_s, y = .resid)) + geom_point(size = 0.1) + geom_smooth(method = "lm", se = FALSE) + labs(x = "Duration (in secs)", y = "Residual", title="Residual of linear model")

#residual is evenly distributed indicating linear model is a good fit for this data
```

```{r}
ggplot(data = spotify.model.df, aes(x = duration_s, y = popularity)) + geom_point(size=0.1) + geom_point(size = 0.1) + geom_smooth(method = "lm", color = "green") + facet_wrap(~ genre, nrow = 1) + labs(x = "Duration (in secs)", y = "Actual Popularity", title="Actual Popularity by Duration for different Genres") + ylim(0,100)
#print(p)
```

```{r}
ggplot(data = spotify.model.df, aes(x = duration_s, y = predictions)) + geom_point(size=0.1) + geom_point(size = 0.1) + geom_smooth(method = "lm", color = "red") + facet_wrap(~ genre, nrow = 1) + labs(x = "Duration (in secs)", y = "Predicted Popularity", title="Predicted Popularity by Duration for different Genres") + ylim(0,100)
```

```{r}
ggplot(spotify.model.df, aes(tempo.level, genre, legend = "Popularity")) + geom_tile(aes(fill = popularity)) + labs(x = "Tempo", y = "Genre", title="Actual Popularity for different combination of Tempo and Genre", fill = "Actual Popularity")
```

```{r}
ggplot(spotify.model.df, aes(tempo.level, genre)) + geom_tile(aes(fill = predictions)) + labs(x = "Tempo", y = "Genre", title="Predicted Popularity for different combination of Tempo and Genre", fill = "Predicted Popularity")
```

```{r}
ggplot(spotify.model.df, aes(speechiness.level, loudness.level)) + geom_tile(aes(fill = popularity))
```

#### Making predictions for songs
```{r}
spotify.df.prediction = spotify.model.df[, ! colnames(spotify.model.df) %in% c("predictions")]
spotify.df.prediction = spotify.df %>% dplyr::select(c(colnames(spotify.df.prediction),"track_name"))
#spotify.df.prediction = spotify.df.prediction[,-spotify.df.prediction$popularity]
spotify.df.prediction$predictions = predict(spotify.lm,  newdata = spotify.df.prediction[,! colnames(spotify.df.prediction) %in% c("predictions","track_name")])

# Creating songs popularity and predictions dataframe
song.df  = spotify.df.prediction[,c("track_name","popularity","predictions")]

write.csv(song.df, "song.predictions.data.csv")
```




### Appendix

## PCA


```{r}
spotify.model.pca.df = spotify.model.df %>% dplyr::select(acousticness, danceability, valence, duration_s)

pca = prcomp(spotify.model.pca.df, scale. = TRUE)
pca.df = as.data.frame(pca$x)

pca.df = cbind(pca.df, spotify.df$genre)
```

```{r}
ggplot(data = pca.df, aes(x = PC1, y = PC2, color=`spotify.df$genre`)) + geom_point() + labs(x = "Principal Component 1", y = "Principal Component 2", title = "Scatterplot of Top 2 Principal Components by Genre")
```


