---
title: "A6_singhvis"
author: "Vishal Singh"
date: "November 30, 2019"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
library(tidyverse)
library(dplyr)
library(tidyverse)
library(lme4)
library(knitr)
library(kableExtra)
library(maps)
library(openintro)
library(ggthemes)
cb_palette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
# Reading Data
countyData = read.csv("county_data.txt",sep = ' ') 

# Selecting required columns
countyData = countyData %>% dplyr::select(name, state, diff_2016, partywinner16, flipped) %>% mutate(state = abbr2state(state))

countyData %>% head() %>% drop_na()
```

```{r}
# Loading county data
us_county = map_data("county")
```

```{r}
# Cleaning the election data
election.df = countyData %>% 
  filter(!state %in% c("Alaska","Hawaii")) %>% 
  mutate(county = str_to_lower(name), state = str_to_lower(state)) %>%
  mutate(county = str_replace_all(county," county| parish|\\.| borough|'| ",'')) %>% 
  drop_na()

us_county = us_county %>% mutate(subregion = str_replace_all(subregion, " ", ""))
election.county.df = election.df %>% inner_join(us_county, by = c('county'='subregion','state' = 'region'))
  
```

### Winning Party by County
When we look at winning parties by counties in the 2016 US Presidential Elections, we see that counties who had democratic winner were mostly concentrated around areas of West Coast, North Eastern states and a few in the states of New Mexico, Colorado and South Texas.

```{r}
# Plotting Election winner by County
ggplot(election.county.df, aes(long, lat, group = group, fill = partywinner16)) + geom_polygon(color = "gray90", size = 0.25) + theme_map() + coord_map(projection = "lambert", lat0 = 30, lat1 = 40) + scale_fill_manual(name = "Party", values = c("dodgerblue3", "indianred3")) + labs(title = "Winning party by County in 2016 US Presidential Election")
```

### Difference in Winning Margin for different counties
When we look at the variable of `Difference in Winning Margin`, we see counties with highest difference in states of California, Oregon for 'Blue States' and Florida, Ohio for 'Red States'.

```{r}
# Plotting difference margin
election.county.df$diff = election.county.df$diff_2016
election.county.df$diff[election.county.df$diff >  10000] = 10000
ggplot(election.county.df, aes(long, lat, group = group, fill = diff)) + geom_polygon(color = "gray90", size = 0.25) + theme_map() + coord_map(projection = "lambert", lat0 = 30, lat1 = 40) + scale_fill_gradient2(name = "Difference of Votes", low = "red", high = "blue")
```

### Counties who "Flipped"
When we look at counties who changed who they voted for from 2012 to 2016, we see most of them concentrated in the North Eastern States and a few in the states of Iowa and Minnesota.

```{r}
# Counties who flipped their party from 2012 to 2016
ggplot(election.county.df, aes(long, lat, group = group, fill = flipped)) + geom_polygon(color = "gray90", size = 0.25) + theme_map() + coord_map(projection = "lambert", lat0 = 30, lat1 = 40) + scale_fill_manual(name = "Flipped", values = c("#999999", "#56B4E9")) + labs(title = "Highlighting Counties who flipped")
```




