---
title: "A5_singhvis"
author: "Vishal Singh"
date: "October 29, 2019"
output: html_document
---


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
library(tidyverse)
library(dplyr)
library(gapminder)
library(tidyverse)
library(lme4)
library(knitr)
library(kableExtra)
library(NHANES)
library(arm)
cb_palette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```


```{r}
nhanes_raw = NHANESraw
nhanes_raw = nhanes_raw  %>% dplyr::select(Diabetes, Age, WTINT2YR) %>% 
  drop_na(Diabetes) %>% 
  mutate(Diabetes = ifelse(Diabetes == "No",0,1))

nhanes = NHANES %>% dplyr::select(Diabetes,Age) %>% 
  drop_na(Diabetes) %>% 
  mutate(Diabetes = ifelse(Diabetes == "No",0,1))

```

## 1
### a.

```{r}
prop_diabetes_nhanes = nhanes %>% group_by(Diabetes) %>% summarise(percent = n()) %>% 
  mutate(percent=percent/sum(percent))

prop_diabetes1a = prop_diabetes_nhanes$percent[prop_diabetes_nhanes$Diabetes == 1]
print(paste0('Proportion of all U.S. residents with diabetes: ', round(prop_diabetes1a,5)))
```

### b.

```{r}
prop_diabetes_nhanesraw = nhanes_raw %>% group_by(Diabetes) %>% summarise(percent = n()) %>% 
  mutate(percent=percent/sum(percent))

prop_diabetes1b = prop_diabetes_nhanesraw$percent[prop_diabetes_nhanesraw$Diabetes == 1]
print(paste0('Proportion of people in the sample with diabetes: ', round(prop_diabetes1b,5)))
```

### c.

```{r}
prop_diabetes_nhanesraw_wt = nhanes_raw %>% group_by(Diabetes) %>% summarise(percent = sum(WTINT2YR)) %>% 
  mutate(percent = percent/sum(percent))

prop_diabetes1c = prop_diabetes_nhanesraw_wt$percent[prop_diabetes_nhanesraw_wt$Diabetes == 1]
print(paste0('Proportion of all U.S. residents with diabetes (weighted): ', round(prop_diabetes1c,5)))
```

The proportion estimated from the data NHANES and NHANESraw (unweighted) is comparable (~8%) and also the proportion estimated from NHANESraw using unweighted and weighted data is also comaparable.

## 2
### a.

Visualize the model:

```{r}
ggplot(nhanes, aes(Age, Diabetes)) + geom_jitter(height = 0.01) + geom_smooth(method = "glm", method.args = list(family = "binomial")) + labs(title ="Unweighted Proportion of U.S. Citizens with Diabetes against Age" ,x ="Age" ,y = "Proportion of U.S citizens with diabetes")
```

Fit a logistic regression:

```{r}
diabetes.logit = glm(Diabetes ~ Age, family = binomial, data = nhanes)
summary(diabetes.logit)
```

**Interpretation** <br>
The coefficient of regression is 0.05634, therefore, we can say that with a unit increase in the value of Age will result in a increase of 0.05634 in the value of log of odds ratio.

### b.

Fit a logistic regression:

```{r}
survey.weights=(nhanes_raw$WTINT2YR/1) 
diabetes.logit.wt = glm(Diabetes ~ Age, family = binomial, data = nhanes_raw, weights = survey.weights)
summary(diabetes.logit.wt)
```

```{r}
#Describing the fit
our.logit = function(x){
coe = coef(diabetes.logit.wt)
y = coe[1] + coe[2] * x
return(exp(y) / (1 + exp(y)))
}
```

Visualize the model:

```{r}
ggplot(nhanes, aes(Age, Diabetes)) + geom_jitter(height = 0.01) + geom_smooth(method = "glm", method.args = list(family = "binomial")) + stat_function(fun = our.logit, color = "orange") + labs(title ="Weighted and Unweighted Proportion of U.S. Citizens with Diabetes against Age" ,x ="Age" ,y = "Proportion of U.S citizens with diabetes")
```

**Interpretation** <br>
The coefficient of regression is 3.192e+13, therefore, we can say that with a unit increase in the value of Age will result in a increase of 3.192e+13 in the value of log of odds ratio. <br>
The coefficients of the two models are very different and the scale of coefficients are not comparable. <br>

## 3

```{r}
nhanes_raw_grp_age_diabetes = nhanes_raw %>% 
  group_by(Age, Diabetes) %>% summarise(wt_sum = sum(WTINT2YR))

nhanes_raw_grp_age = nhanes_raw %>% 
  group_by(Age) %>% summarise(wt_sum = sum(WTINT2YR))

nhanes_prop = nhanes_raw_grp_age_diabetes %>%  left_join(nhanes_raw_grp_age, by = 'Age') %>% 
  mutate(age_diabetes_prop = wt_sum.x/wt_sum.y) %>% filter(Diabetes == 1)
```

```{r}
ggplot(nhanes_prop, aes(Age, age_diabetes_prop)) + geom_point() +
  stat_function(fun = our.logit, color = "orange") + labs(title ="Proportion of people with diabetes by Age and Logit model" ,x ="Age" ,y = "Proportion of U.S citizens with diabetes")
```

The curve for the proportions for people who have diabetes does not match with the curve for logits using weights. <br>
The weights for people which do not have diabetes is very high compared to people who have diabetes, therefore, our model only predicts 0.

